% ============================================================
%  SUPPLEMENTARY MATERIALS  —  paste as a separate document
%  or \input{supplementary} after \appendix
%
%  Requires: amsmath, booktabs, array, xspace, hyperref (optional)
%  Compile with pdflatex or lualatex.
% ============================================================

\section*{Supplementary Materials}
\addcontentsline{toc}{section}{Supplementary Materials}

% Renumber tables and figures with S prefix
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{equation}{0}

% ============================================================
\subsection*{S1\quad Model Parameters}

\begin{table}[h]
\centering
\caption{Fixed simulation parameters (all experiments).}
\label{tab:S1}
\small
\begin{tabular}{llll}
\toprule
Parameter & Symbol & Value & Rationale \\
\midrule
Grid width / height         & ---  & $30\times30$   & Sufficient spatial resolution; tractable for $N=20$ replications \\
Number of human agents      & $N$  & 100            & Large enough for network effects; small enough to track individually \\
AI agents                   & ---  & 5              & $\approx$1 AI per 20 humans; non-negligible but not dominant coverage \\
AI sensing fraction (per tick) & ---& 0.15          & AI observes 15\% of the 900-cell grid ($\approx$135 cells/tick) \\
Share of exploitative agents & ---  & 0.50          & Balanced population; varied in factor sweep (Table~\ref{tab:S8}) \\
Share of confirming social ties & --- & 0.70        & Reflects documented homophily in disaster communication \\
Disaster dynamics mode      & ---  & 2              & Medium-tempo: cells drift $\pm1$--$2$ levels/tick toward baseline \\
Shock probability (per cell/tick) & --- & 0.10      & Persistent environmental non-stationarity \\
Shock magnitude             & ---  & $\pm2$ levels  & Produces salient deviations requiring active information-seeking \\
Simulation length           & $T$  & 200 ticks      & Allows learning, echo-chamber formation, and steady-state measurement \\
Initial (human) trust       & ---  & 0.30           & Low baseline prevents unrealistic blind trust at $t=0$ \\
Initial AI trust            & ---  & 0.25           & Slight AI scepticism, consistent with survey findings \\
Q-learning rate             & $\eta$ & 0.10         & Standard value; sensitivity analysis unreported \\
$\varepsilon$-greedy exploration & $\varepsilon$ & 0.30 & Ensures continued source diversity throughout run \\
Trust learning rate (exploitative) & --- & 0.015   & Slow trust revision matches confirmation-seeking profile \\
Trust learning rate (exploratory)  & --- & 0.030   & Faster but still sub-linear trust revision \\
Rumour probability          & ---  & 1.00           & All social information propagated (worst-case echo scenario) \\
Relief outcome delay        & ---  & 15--25 ticks   & Random; models logistical pipeline in humanitarian response \\
Steady-state window         & ---  & 75 ticks       & Last 75 of 200 ticks (15 samples at the 5-tick cadence) \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\subsection*{S2\quad Agent Cognitive Parameters}

\begin{table}[h]
\centering
\caption{Bayesian acceptance parameters by agent type.}
\label{tab:S2}
\small
\begin{tabular}{llll}
\toprule
Parameter & Exploitative & Exploratory & Interpretation \\
\midrule
Acceptance half-width, $D$   & 2.0 & 4.0 & Distance at which $P(\text{accept})=0.5$ \\
Acceptance steepness, $\delta$ & 3.5 & 1.2 & Higher $\delta$ $\Rightarrow$ sharper threshold \\
Precision weight             & $1.5\times c/(1-c)$ & $0.8\times c/(1-c)$ & Resistance to belief revision ($c$ = confidence) \\
Sensing radius               & 0 cells & 0 cells & Both types rely on communication \\
Interest-point targeting     & Believed epicentre & Highest-uncertainty cell & Governs query direction \\
\bottomrule
\end{tabular}
\end{table}

The acceptance probability for report $r$ given prior belief $b$ is
%
\begin{equation}
  P(\text{accept}) = \frac{D_{\mathrm{eff}}^{\delta}}{|r-b|^{\delta}+D_{\mathrm{eff}}^{\delta}},
  \tag{S1}
\end{equation}
%
where $D_{\mathrm{eff}} = D\cdot(1+0.5\,T_{\mathrm{source}})$ for social
contacts and $D_{\mathrm{eff}}=D$ for the AI.
This generalises the bounded-confidence model of \citet{Deffuant2000} by
replacing the binary threshold with a smooth curve parameterised by $(D,\delta)$.

% ============================================================
\subsection*{S3\quad AI Alignment Mechanism}

The AI response formula is
%
\begin{equation}
  r_{\mathrm{AI}} = (1-\alpha)\,t + \alpha\,b,
  \tag{S2}
\end{equation}
%
where $t$ is the ground-truth cell severity, $b$ is the querying agent's current
belief, and $\alpha\in[0,1]$ is the alignment parameter.
At $\alpha=0$: $r_{\mathrm{AI}}=t$ (fully truthful).
At $\alpha=1$: $r_{\mathrm{AI}}=b$ (perfect confirmation).
Intermediate values produce a weighted average --- the AI is partially informative
and partially sycophantic.

This linear interpolation is the simplest formulation spanning the
truth--confirmation spectrum without additional free parameters.
Non-linear or stochastic variants (e.g.\
$r\sim\mathcal{N}(\alpha b+(1-\alpha)t,\,\sigma^2)$) were not pursued in the
main sweep to preserve interpretability; adding symmetric noise shifts all
SECI/AECI curves uniformly without displacing $\alpha^*$.

% ============================================================
\subsection*{S4\quad Cognitive Gap Scalar}

The gap scalar $g$ parameterises cognitive divergence between agent types around
a shared midpoint:
%
\begin{align}
  D_{\mathrm{exploit}}(g) &= \max(3.0 - 1.0\,g,\;0.1),
  &\delta_{\mathrm{exploit}}(g) &= 2.35 + 1.15\,g, \tag{S3}\\
  D_{\mathrm{explor}}(g)  &= 3.0 + 1.0\,g,
  &\delta_{\mathrm{explor}}(g)  &= \max(2.35 - 1.15\,g,\;0.1). \tag{S4}
\end{align}

\begin{table}[h]
\centering
\caption{Cognitive parameters at each gap-scalar value.}
\label{tab:S4}
\begin{tabular}{ccccc}
\toprule
$g$ & $D_{\mathrm{exploit}}$ & $\delta_{\mathrm{exploit}}$ & $D_{\mathrm{explor}}$ & $\delta_{\mathrm{explor}}$ \\
\midrule
0.0 & 3.00 & 2.35 & 3.00 & 2.35 \\
0.5 & 2.50 & 2.93 & 3.50 & 1.78 \\
1.0 & 2.00 & 3.50 & 4.00 & 1.20 \\
1.5 & 1.50 & 4.08 & 4.50 & 0.63 \\
\bottomrule
\end{tabular}
\end{table}

The shared midpoint ($D_{\mathrm{mid}}=3.0$, $\delta_{\mathrm{mid}}=2.35$) is
the arithmetic mean of the baseline exploit/explor values.
At $g=0$ both types are cognitively identical, providing a null condition for
heterogeneity effects.
The invariant $D_{\mathrm{exploit}}<D_{\mathrm{explor}}$ and
$\delta_{\mathrm{exploit}}>\delta_{\mathrm{explor}}$ holds for all $g>0$,
ensuring the exploitative type always has a narrower and steeper acceptance curve.

The gap sweep extends the design to a second axis: $d_{\mathrm{mid}}\in\{2.0,3.0,4.0\}$
independently varies the absolute level of cognitive openness (closed, baseline, open)
orthogonally to the inter-type gap $g$.
The full 2D sweep uses $N=20$ independently seeded replications per $(g,d_{\mathrm{mid}},\alpha)$
cell and $T=200$ ticks, matching the primary alignment sweep.

% ============================================================
\subsection*{S5\quad Metric Definitions}

\subsubsection*{S5.1\enspace Social Echo Chamber Index (SECI)}

\begin{equation}
  \mathrm{SECI} = 1 -
    \frac{\mathrm{Var}\!\bigl(\{b_{f,k}:f\in\mathrm{friends}(i),\,k\in\mathcal{K}\}\bigr)}
         {\mathrm{Var}\!\bigl(\{b_{j,k}:j\in\mathcal{A},\,k\in\mathcal{K}\}\bigr)},
  \tag{S5}
\end{equation}
%
where $\mathcal{A}$ is the set of all agents and $\mathcal{K}$ is the set of all
grid cells.
Computed separately for exploitative and exploratory agents every five ticks.
A \emph{formation} event is registered when $\mathrm{SECI}<-0.30$ sustained for
$\geq3$ consecutive samples; a \emph{recovery} event when
$\mathrm{SECI}\geq0.0$ sustained for $\geq5$ consecutive samples after formation.
If formation never occurs the lifecycle bar is left blank ($\circ$); if formation
occurs but recovery does not before the run ends the bar reaches maximum height
with annotation $\times$.

\subsubsection*{S5.2\enspace AI Echo Chamber Index (AECI)}

\begin{equation}
  \mathrm{AECI} = \frac{\Delta q_{\mathrm{AI}}}
                       {\Delta q_{\mathrm{AI}} + \Delta q_{\mathrm{human}}},
  \tag{S6}
\end{equation}
%
where $\Delta q$ denotes the increment in \emph{accepted} information-update
calls over a five-tick window.
This window-delta formulation distinguishes recent reliance from cumulative
history, making AECI responsive to within-run behavioural shifts.

\subsubsection*{S5.3\enspace Belief Accuracy (MAE)}

\begin{equation}
  \mathrm{MAE} = \frac{1}{|\mathcal{C}|}\sum_{k\in\mathcal{C}}|b_{i,k}-t_k|,
  \tag{S7}
\end{equation}
%
averaged over agents $i$, where $\mathcal{C}=\{k:t_k\geq1\}$ is the set of
cells with non-zero true severity.

\subsubsection*{S5.4\enspace Composite Goldilocks Scores}

Let $\bar{m}(\alpha)$ denote the steady-state mean (last 75 ticks) of metric $m$
at alignment level $\alpha$.
Range normalisation across the sweep:
%
\begin{equation}
  m_{\mathrm{norm}}(\alpha) =
    \frac{\bar{m}(\alpha)-\min_{\alpha}\bar{m}}
         {\max_{\alpha}\bar{m}-\min_{\alpha}\bar{m}}.
  \tag{S8}
\end{equation}
%
Composite scores:
%
\begin{align}
  \text{total\_bubble}_{\mathrm{norm}}(\alpha)
    &= |\mathrm{SECI}|_{\mathrm{norm}}(\alpha)
       +|\mathrm{AECI}|_{\mathrm{norm}}(\alpha), \tag{S9}\\
  \text{total\_score}_{\mathrm{norm}}(\alpha)
    &= |\mathrm{SECI}|_{\mathrm{norm}}(\alpha)
       +|\mathrm{AECI}|_{\mathrm{norm}}(\alpha)
       +\mathrm{MAE}_{\mathrm{norm}}(\alpha). \tag{S10}
\end{align}
%
Optima:
%
\begin{align}
  \alpha^* &= \arg\min_{\alpha}\,\text{total\_bubble}_{\mathrm{norm}}(\alpha), \tag{S11}\\
  \alpha^*(+\mathrm{MAE}) &= \arg\min_{\alpha}\,\text{total\_score}_{\mathrm{norm}}(\alpha). \tag{S12}
\end{align}
%
Range normalisation ensures each component contributes on the same $[0,1]$ scale
regardless of the absolute magnitude of SECI, AECI, or MAE.
The two optima are reported separately so that the trade-off between
echo-chamber suppression and accuracy improvement is explicit.

\subsubsection*{S5.5\enspace Spatial and Network Periphery Metrics}

\paragraph{Spatial periphery.}
For each simulation run, cells are ranked by Euclidean distance to that run's
epicentre.
\emph{Near} cells (Q1, closest quartile) and \emph{far} cells (Q4, furthest
quartile) are identified per run; coverage deficit (actual severity minus mean
aid tokens per tick) and aid density are averaged within each set and then
averaged across replications.
This per-run computation avoids the blurring artefact that would arise from
projecting all replications onto a single grid when epicentre locations vary.

\paragraph{Network periphery.}
Agents are ranked by social-network degree.
Low-degree agents (Q1) and high-degree agents (Q4) are compared on belief MAE
and relief tokens sent per tick.

% ============================================================
\subsection*{S6\quad Primary Experiment: Goldilocks Alignment Sweep}

\begin{table}[h]
\centering
\caption{Experiment design: primary Goldilocks alignment sweep.}
\label{tab:S6}
\begin{tabular}{lll}
\toprule
Dimension & Values & $N$ per cell \\
\midrule
AI alignment $\alpha$
  & $0.0,\,0.1,\,0.2,\ldots,1.0$ (11 levels) & 20 \\
Agent type mix
  & 50\% exploit\,/\,50\% explor (fixed) & --- \\
Simulation length $T$
  & 200 ticks & --- \\
\midrule
\textbf{Total simulation runs} & \multicolumn{2}{l}{\textbf{220}} \\
\bottomrule
\end{tabular}
\end{table}

Each replication uses an independent random seed for epicentre placement, agent
initialisation, and shock sequences.
Results are presented as boxplots (median, IQR, $1.5\times$IQR whiskers) over the
20 replications.
Steady-state scalars are averaged over the last 75 ticks (15 samples at the
5-tick cadence).

% ============================================================
\subsection*{S7\quad Secondary Experiment: Cognitive Gap Sweep}

\begin{table}[h]
\centering
\caption{Experiment design: cognitive gap sweep.}
\label{tab:S7}
\begin{tabular}{lll}
\toprule
Dimension & Values & $N$ per cell \\
\midrule
Gap scalar $g$               & $\{0.0,\,0.5,\,1.0,\,1.5\}$                   & 20 \\
Acceptance midpoint $d_{\mathrm{mid}}$ & $\{2.0\ (\text{closed}),\ 3.0\ (\text{baseline}),\ 4.0\ (\text{open})\}$ & --- \\
AI alignment $\alpha$ (within each $g$, $d_{\mathrm{mid}}$)
                             & $0.0,\,0.1,\ldots,1.0$ (11 levels)             & 20 \\
Simulation length $T$        & 200 ticks                                       & --- \\
\midrule
\textbf{Total simulation runs} & \multicolumn{2}{l}{\textbf{2640} ($4\times3\times11\times20$)} \\
\bottomrule
\end{tabular}
\end{table}

Each replication uses an independent random seed (seed\,=\,run index).
The $d_{\mathrm{mid}}$ axis tests whether the Goldilocks finding is robust to
the absolute level of cognitive openness, independently of the inter-type gap $g$.
Results confirm $\alpha^*=0.8$ across all 132 $(g,d_{\mathrm{mid}})$ conditions;
the full $N=20$ design is treated as confirmatory.
Both $\alpha^*(\text{bubble})$ and $\alpha^*(+\mathrm{MAE})$ are shown in the
figure (solid and dashed lines respectively).

% ============================================================
\subsection*{S8\quad Factor Sensitivity Sweeps}

Three one-at-a-time factor sweeps test the robustness of the Goldilocks finding.
All use $\alpha=0.5$ and $N=2$ replications with run length
$\max(50,T/2)=125$ ticks.

\begin{table}[h]
\centering
\caption{Factor sensitivity sweep design.}
\label{tab:S8}
\begin{tabular}{lll}
\toprule
Factor & Values swept & Fixed parameters \\
\midrule
Share of exploitative agents & $\{0.2,\,0.5,\,0.8\}$
  & All base parameters except \texttt{share\_exploitative} \\
Disaster dynamics mode & $\{0,\,2,\,3\}$
  & All base parameters except \texttt{disaster\_dynamics} \\
Rumour probability & $\{0.0,\,0.5,\,1.0\}$
  & All base parameters except \texttt{rumor\_probability} \\
\bottomrule
\end{tabular}
\end{table}

Disaster dynamics mode controls the rate at which cell severities drift and
receive shocks: mode~0 fixes the initial map; mode~2 applies the baseline shock
schedule (Table~\ref{tab:S1}); mode~3 doubles shock frequency.

% ============================================================
\subsection*{S9\quad Software and Reproducibility}

The model is implemented in Python~3.11 using Mesa~3.x for agent scheduling,
NetworkX for social-network construction, and NumPy / Matplotlib for numerical
computation and visualisation.
All simulation runs are seeded via \texttt{numpy.random.seed(run\_index)},
ensuring bit-for-bit reproducibility given the same Python and library versions.
Results are serialised to JSON after each $(\alpha,\mathrm{run})$ cell; the
plotting pipeline reads these files without re-running the simulation, so figures
can be regenerated without recomputation.

\begin{table}[h]
\centering
\caption{Software dependencies.}
\label{tab:S9}
\begin{tabular}{ll}
\toprule
Package & Version \\
\midrule
Python     & 3.11 \\
Mesa       & $\geq$3.0 \\
NetworkX   & $\geq$3.0 \\
NumPy      & $\geq$1.24 \\
Matplotlib & $\geq$3.7 \\
SciPy      & $\geq$1.10 \\
\bottomrule
\end{tabular}
\end{table}

The full codebase is available at the repository listed on the title page.
Raw JSON result files for the main sweep ($N=20$) require approximately 150~MB
of disk space and are archived separately.
